PanLex

### 2013-01-01 paper/2012/SWJ-PanLex/paper.tex
Added Jonathan's suggestions

\begin{abstract}
At present, there are approximately 6,900 actively spoken languages in the world.
However, some experts claim, that the process of globalization may eventually lead
to a world loosing this linguistic diversity.
The vision of the PanLex project is to save these languages, especially low density ones,
by allowing them to be a part of the Information Age by making them applicable via translations.
For this reason, PanLex gathers and integrates information from several
linguistict sources, such as monolingual dictionaries, bilingual dictionaries, multilingual dictionaries,
glossaries, standards and thesauri.
In this paper we describe how we transformed this data to RDF, interlinked it with
lexvo and DBpedia and published it as Linked Data and SPARQL.
\end{abstract}

@@

\section{Introduction}
% \begin{itemize}
% \item introduce PanLex project
% \item @author: Jonathan
% \item Motivation: About approximately 6900 active languages (\todo{source seems to be http://www.amazon.com/Ethnologue-Languages-M-Paul-Lewis/dp/1556712162}), Existing translation only between relatively few language pairs (dictionaries).
% \item Transitive word-by-word translations 
% * high density and low density languages
% * decline in linguistic diversity
% * These combined forces have led to predictions that something between half and 90%
%   of the world’s living languages will die within the next century (Woodbury 2006; UNESCO 2003: 2).
%   Weaker forms of these forces appear to be shrinking the use of medium-density languages in science,
%   diplomacy, business, and other domains (Phillipson 2008).
% * When languages die, they argue, the world loses
%   (1) irreplaceable knowledge of history, medicine, nature, and productive methods encoded in languages’
% * extraction + integration
% \end{itemize}
At present, there are about 7,000 languages in the world with approximately 6,900 of them being actively spoken\footnote{see \url{http://www.sil.org/iso639-3/download.asp} for a list of registered languages}.
Nonetheless, some experts claim, that the process of globalization may lead to a monolingual world loosing this linguistic diversity~\cite{endang_lang}.
Since there are a lot of low density languages that were not taken over from generation to generation anymore, we could irreversibly loose parts of human knowledge.
Some predictions state that this decline may lead to a loss of half to 90\% of human's languages~\cite{endang_lang}.

The vision of the PanLex project is to save all these thousands of languages, especially those low-density ones that are threatened by extinction.
PanLex tries to provide means for panlingual translations from any language of the world into any other.
But it is not feasible to just store all the translations of one word into any other language since this would result in trillions of translations.
So the PanLex project additionally uses artificial intelligence techniques to infer translations.
As a consequence, not just mainstream languages like English, but even these low-density languages may be used for a global communication.
% Since it is not feasible to just store all the translations of one word into any other language, which would result in trillions of translations, a system is needed that allows translations by inference using artificial intelligence.

Currently there is a growing community trying to combine linguistic knowledge with Semantic Web technologies and so build a web of linguistic data, also known as the \emph{Linguistic Linked Open Data (LLOD) cloud}.
Originally the PanLex project provides different web services\footnote{\url{http://panlex.org/try/}} and APIs\footnote{\url{http://panlex.org/tech/doc/api/}} to access its data.
Since the project shares the idea of open access we tried to make PanLex a part of the LLOD cloud.

@@

\section{Triplification of the Raw Data}
\label{sec:triplificate}
In this section we first provide an analysis of the original PanLex data.
Subsequently, we will explain the steps taken to publish the data as RDF.
The vocabulary introduced for the PanLex dataset will be described and afterwards we will show how the original data was converted to RDF and published via a SPARQL endpoint.
The interlinking of the created RDF data with other datasets will be discussed in the following section.

\subsection{Analysis of the Original Dataset}
\label{sec:analysis}
At the core, PanLex is a \emph{database}, served by a PostgreSQL database server, that ``represents \emph{assertions} about the \emph{meanings} of \emph{expressions}''\footnote{\url{http://panlex.org/tech/doc/design/panlex-db-design.pdf}}.
The database contains about 20 million meanings and 19 million expressions extracted from about 4.000 sources.

Assertions are factual claims attributed to a certain source and editor.
They do not represent absolute truths and as such it is valid that different editors make contradictory assertions on different sources.

@@

The lexical database of the PanLex project is based on different sources like monolingual dictionaries,
bilingual dictionaries, multilingual dictionaries, glossaries, standards, thesauri and so on.
The list of used sources can also be found on the web\footnote{\url{http://panlex.org/tech/plrefs.shtml}}.

@@

\begin{itemize}
  \item \emph{Expressions} are lemmas. For example, \emph{``go''} is a valid expression, whereas \emph{``went''} is not.
  Expressions are always given in a language variety and can only be given once per language variety.
  They were extracted from the above mentioned mono- or bilingual dictionaries or from other sources.
  \item \emph{Languages} in PanLex are identified using ISO 639-3\footnote{\url{http://www.sil.org/iso639-3/codes.asp}} individual and macrolanguage codes, ISO 639-2\footnote{\url{http://www.loc.gov/standards/iso639-2/}} collective codes and ISO 639-5\footnote{\url{http://www.loc.gov/standards/iso639-5/}} codes.
  \item \emph {Language Varieties} within a language are further distinguished by custom names.
  For example, ``eng'' is the ISO 639-3 code for the English language.
  Panlex defines ``simple english'' and ``british english'' as varieties of it.
  These can be given in the native writing system.
  So, in contrast to mechanisms like IETF BCP 47\footnote{\url{http://tools.ietf.org/html/bcp47}} there is no need for a transcription. %  and one can use variety names like ``{زبان اوستایی}''.
  Additionally these variety names do not have to be registered like the IETF BCP 47 language subtags\footnote{\url{http://www.iana.org/assignments/language-subtag-registry}}.
  \item \emph{Meanings} of expressions are stated by users based on their
  interpretation of an information source.
  Information sources can be dictionaries, thesauri but also personal knowledge. 
  The pair comprised of the user and the information source is referred to as a meaning's \emph{approver}.
  Users may add informal definitions to meanings in any language variety.
  \item \emph{Definitions} are optional descriptions of a meaning.
  They are given in a certain language variety.
  A description of the verb \emph{browse} for example could be \emph{``move or surf through various files on a computer, the Internet, etc.''}.
  \item \emph{Denotations} relate expressions to possible \emph{meanings}.
  So homonyms are represented by one expression being connected to different meanings.
  Denotations may be enriched with parts of speech tags adopted from the
  Open Lexicon Interchange Format (OLIF) standard\footnote{\url{http://www.olif.net/}}.
  For example, \emph{fall} can be a verb or a noun for autumn.
  Furthermore, it is possible to generically tag denotations with meta data
  using key-value pairs.
  For instance, an English expression \emph{pig} when synonymous with
  \emph{police officer} could be annotated with \emph{pragmatics=vulgar}.
  \item \emph{Users} are related to their defined language varieties and edited approvers.
  \item \emph{Licenses} are also considered by the PanLex project.
  At present there are ten different license categories an approver can get into.
  They are \emph{public domain}, \emph{Creative Commons}, \emph{request} (meaning that one has to ask the author of the source at hand), \emph{GNU General Public License}, \emph{GNU Lesser General Public License}, \emph{GNU Free Documentation License}, \emph{MIT License}, \emph{copyright} (stating that there is a certain copyright holder), \emph{other} and \emph{unknown}.
\end{itemize}

An overview of the number of instances per entity in the PanLex database is given in~\autoref{fig:plx-entity-counts}.
\begin{table}
\centering
\begin{scriptsize}
\begin{tabular}{lr}
\toprule
Entity & Instances \\
%\midrule
Expressions & 18,580,594 \\ % SELECT COUNT(ex) FROM ex;
%\midrule
Languages   & 7,839 \\ % SELECT COUNT(lc) FROM lc; -- all language codes (ISO 639-3, ISO 639-2, ISO 639-5) known in the PanLex db
% FIXME: there should not be more languages than language varieties
%\midrule
Language Varieties & 7,248 \\ % SELECT COUNT(lv) FROM lv;
%\midrule
Meanings    & 20,023,427 \\ % SELECT COUNT(mn) FROM mn;
%\midrule
Definitions & 2,522,605 \\ % SELECT COUNT(df) FROM df;
%\midrule
Denotations & 50,803,243 \\ % SELECT COUNT(dn) FROM dn;
%\midrule
Users       & 7 \\ % SELECT COUNT(us) FROM us;
%\midrule
Licenses    & 10 \\ % SELECT COUNT(li) FROM apli;
\bottomrule
\end{tabular}
\end{scriptsize}
\caption{Number of entities in the PanLex database}
\label{fig:plx-entity-counts}
\end{table}

@@

There are two noteworthy limitations of the database schema:
Approver combines a user with an information source, however the \emph{information source} is not modelled as a distinct entity.
Users are only associated with language variety and approver entities. However,
it is not possible to track which denotations were added by users.
\todo{What are the consequences? }

@@

The RDF analogon of the PanLex \emph{meaning} is the \texttt{plx:Meaning}.
Entities of this class may have an identifier assigned with the \texttt{plx:identifier} property pointing to an \texttt{xsd:string} literal.
This meaning identifier represents ids given to the considered meaning in other data sources and may be used to build links to the meaning as represented on these sites.
Meanings may also have \emph{definitions}, entities of the \texttt{plx:Definition} class, giving a textual representation (\texttt{rdfs:label}) with more than just a lexeme in a certain language variety (\texttt{plx:languageVariety}).

@@

\subsection{Transformation workflow}
\label{sec:conversion}
\todo{how often in PanLex updated?}
The size of the PanLex database (including indexes) is approximately 18 GB,
which is big enough to make recurrent RDF conversion cumbersome:
A full RDF conversion comes bloats in the dataset size and data duplication.   
Using conventional hardware, it takes quite long
to convert all the data, which usually makes
testing and debugging and fixing issues in the modelling or data conversion of the
data to become very time consuming.
As the PanLex data already resides in a relation database, the use of a virtual
RDB 2 RDF mapping solution is a natural choice. The \emph{Sparqlify Mapping Language} thereby
offers a very easy to use mapping language.
Essentially, these mappings consist of three clauses:
\begin{itemize}
  \item \emph{From} Specifies the logical SQL table (i.e. table, view or query) underlying the Sparqlify-ML view.
  \item \emph{With} Defines SPARQL variables by means of expressions over relational columns that yield RDF terms.
  \item \emph{Construct} The template (i.e. the set of triple patterns) to be generated in this view, based on the SPARQL variable definitions.
\end{itemize}
~\autoref{fig:ex:sparqlify-ml} shows an example of a Sparqlify-ML view for the
languages in PanLex: From each row of the \emph{i1} table three resources are
bound to the variable names \emph{?lang}, \emph{?iso3} and \emph{?lexvo3}.
Resources for \emph{?lang} become typed as a \emph{Language}in the PanLex and the \emph{schema.org} namespace.
\todo{Maybe we should link to lexvo directly.}
This view based approach also demonstrates that changing the vocubulary or
adding support for now ones do no require an ETL process, and can therefore be done with little efforts.
%PanLex's languages  iso 639-3 code 
%typed with plx:Language
%resources are created for
%the languages in PanLex, linked with \emph{lexvo}, and

@@

Since Lexvo covers all ISO 639-3 language codes, interlinking could be done with
the Sparqlify-ML view in Figure~\ref{fig:ex:sparqlify-ml}.
For DBpedia we were interested in creating \emph{valid} and
thus \emph{dereferenceable} links.
Therefore, we iterated the \emph{titles} datasets\footnote{\url{http://wiki.dbpedia.org/Downloads38} - Note that more languages are available in the raw download section at \url{http://downloads.dbpedia.org/3.8/}},
which map non-localized DBpedia URIs to their page titles in the respective language.
For each language version we normalized the labels in the same way as being done in PanLex, i.e.
Unicode NFKD\footnote{\url{http://unicode.org/reports/tr15/}} normalization and
removal of punctuation characters are applied.
The DBpedia resources are then mapped to the PanLex expression that are equal to the resource's normalized label in the respective language.
Table ~\ref{fig:plx-dbp-link-counts} summarizes the number of links obtained.
\begin{table}
\centering
\begin{scriptsize}
\begin{tabular}{llll}
\toprule
Language & 639-1 & 639-3 & Links \\
\midrule
English    & en & eng & 1,415,241 \\
German     & de & deu &   224,146 \\
French     & fr & fra &   187,364 \\
Italian    & it & ita &   147,485 \\
Spanish    & sp & spa &   117,056 \\
Portuguese & pt & por &   112,266 \\
Polish     & pl & pol &   110,974 \\
Russian    & ru & rus &    68,040 \\
Czech      & cs & ces &    28,767 \\
Catalan    & ca & cat &    27,779 \\
Korean     & ko & kor &    24,912 \\
Turkish    & tr & tur &    22,258 \\
Bulgarian  & bg & bul &    19,431 \\
Hungarian  & hu & hun &    18,203 \\
Redo Slovene & sl & slv &  4,636 \\
Greek      & el & ell &  1,112 \\
\bottomrule
\end{tabular}
\end{scriptsize}
\caption{Number of DBpedia links per language}
\label{fig:plx-dbp-link-counts}
\end{table}

@@

With the RDF version of PanLex, a large lexical resource participates in the Web of Data,
making it available to a broader community. Users can easily explore the PanLex data using
the services provided by SNORQL and Pubby. This also increases the chance of development of
interesting mesh ups.
For example, the TeraDict translation service\footnote{\url{http://panlex.org/cgi-bin/apidemo-2.cgi}}
could be easily realized using simple SPARQL query as shown in
\todo{Create the query: translations in a given target language of an expression}. 
Anothe future usage scenario is to link PanLex to Wortschatz in order to 
improve the recall of word-by-word translations by considering co-occurences.

\section{Related Work}
\label{sec:related}
PanLex is an integration project of many existing lexical resources with
the goal of eventually covering all languages in the world, especially
ones having low numbers in their speakers.
As such, the extraction of information from linguistic sources, as well as
techniques for automatically inferring translations are
relevant work discussed in \cite{panlex_probtrans}.

An important initiative is the
\emph{Global Wordnet Association}\footnote{\url{http://www.globalwordnet.org/}},
which offers a platform for sharing word nets and defines several goals.
These include setting forth standards for uniformly representing wordnets
of different languages and establishing an universal index of meaning.

@@

At the level of the Semantic Web, there is currently a trend in publishing linguistic resources
as Linked Data\footnote{\url{http://www.w3.org/DesignIssues/LinkedData.html}} and via SPARQL{\footnote{\url{http://www.w3.org/TR/sparql11-query/}}}.
These efforts are referred to as the \emph{Linguistic Linked Open Data Cloud} (LLOD)\footnote{\url{http://linguistics.okfn.org/resources/llod/}}. 
%Some of these datasets already make use of quasi-standard linguistic Semantic Web Ontolgies, such as  
Thereby, several (quasi-)standard ontologies have been developed for covering different aspects of linguistic resources.
Examples include the \emph{Ontologies of Linguistic Annotation} (OLiA)~\cite{olia2010},
the \emph{Lexicon Model for Ontologies} (lemon)~\cite{lemon2011} for modeling lexicon and machine-readable dictionaries,
\emph{POWLA} for modeling linguistic corpora\cite{powla2012}
and the \emph{Natural Language Processing Interchange Format} (NIF)\footnote{\url{http://nlp2rdf.org/nif-1-0}}.

@@

\section{Conclusions and Future Work}
\label{sec:conclusion}
In this paper we describe the PanLex database and its conversion to RDF.
Based on our URI and vocabulary design, we created appropriate view definitions
for the virtual RDB to RDF solution \emph{Sparqlify} which carries out
the actualy RDF transformation.
Furthermore, we interlinked the languages in PanLex with lexvo,
and created a total of \todo {number} derefenceable
links to DBpedia for expressions in 16 languages.
%We detailed the developed vocabulary and

In this work we also identified some weaknesses in
the orginal database design,
which we intend to overcome in the future:
The orginal dataset currently does not cleanly model information sources and approvers
as distinct entities.\todo{We should still map it like that with Sparqlify - resuse ap for the source}
Also, while it is possible to determine
a meaning's source by its approver, this is not possible for the denotations.
Retaining this information seems beneficial for translation approches,
as this enables for example the exclusion of a poor quality or for a certain task
inappropriate source.

>>>>

\begin{abstract}
At present, there are approximately 7,000 living languages in the world.
However, some experts claim that the process of globalization may eventually lead
to the world losing this linguistic diversity.
The vision of the PanLex project is to save these languages, especially low-density ones,
by allowing them to be intertranslatable and thus to be a part of the Information Age.
For this reason, PanLex gathers and integrates information from thousands of
linguistic resources, such as monolingual dictionaries, bilingual dictionaries, multilingual dictionaries,
glossaries, standards and thesauri.
In this paper we describe how we transformed this data to RDF, interlinked it with
lexvo and DBpedia and published it as Linked Data and SPARQL.
\end{abstract}

@@

\section{Introduction}
% \begin{itemize}
% \item introduce PanLex project
% \item @author: Jonathan
% \item Motivation: About approximately 6900 active languages (\todo{source seems to be http://www.amazon.com/Ethnologue-Languages-M-Paul-Lewis/dp/1556712162}), Existing translation only between relatively few language pairs (dictionaries).
% \item Transitive word-by-word translations 
% * high density and low density languages
% * decline in linguistic diversity
% * These combined forces have led to predictions that something between half and 90%
%   of the world's living languages will die within the next century (Woodbury 2006; UNESCO 2003: 2).
%   Weaker forms of these forces appear to be shrinking the use of medium-density languages in science,
%   diplomacy, business, and other domains (Phillipson 2008).
% * When languages die, they argue, the world loses
%   (1) irreplaceable knowledge of history, medicine, nature, and productive methods encoded in languages'
% * extraction + integration
% \end{itemize}
At present, there are about 7,000 living languages in the world\footnote{See \url{http://www.sil.org/iso639-3/download.asp} for a list of registered languages}.
Nonetheless, some experts claim that processes such as nation-state consolidation and globalization are producing language attrition so rapidly that up to 90\% of all languages alive today will be extinct within a century~\cite{endang_lang}.
Theorists of biolinguistic diversity~\cite{nettle} argue that the loss of
language diversity, the loss of human biological knowledge, and the loss of
species diversity are mutually supportive and thus that language preservation
and revitalization are essential to the preservation of biological diversity.
The vision of the PanLex project is to save all these thousands of languages,
especially those low-density ones that are threatened by extinction,
by supporting their use in global communication. This requires panlingual
translation: translation from any language of the world into any other.
One of the crucial components of panlingual translation of discourses is
panlingual lexical translation. PanLex is designed to support that component.
It documents the known lexical translations (translations of lexemes) among all
languages.

Although a list of all the translations of all words into all other languages
would be large (amounting to trillions of translations), there are more
serious obstacles. The vast majority of these translations are unknown.
Those that are known are not universally agreed on. And lexical polysemy
and ambiguity make the question ``What is the translation of lexeme X into language Y?''
underspecified for practical purposes.
So the PanLex project is designed to permit the application of artificial
intelligence techniques to infer and select translations.

Currently there is a growing community trying to combine linguistic knowledge
with Semantic Web technologies and thereby build a Web of Linguistic Data,
also known as the \emph{Linguistic Linked Open Data (LLOD) cloud}.
The PanLex project is still in its data-acquisition phase, but it has provided a
few web services\footnote{\url{http://panlex.org/try/}} and
APIs\footnote{\url{http://panlex.org/tech/doc/api/}} to access its data.
Since the project shares the idea of open access, we have undertaken the project
of making PanLex a part of the LLOD cloud.

@@

\section{Triplification of the Raw Data}
\label{sec:triplificate}
In this section we first provide an analysis of the original PanLex data.
Subsequently, we introduce our URI and vocabulary design and we explain 
the steps taken to publish the data as RDF.
%The vocabulary 
%afterwards we show how the original data was converted to RDF and published
%via a SPARQL endpoint.
%The interlinking of the created RDF data with other datasets is discussed in the
%following section.

\subsection{Analysis of the Original Dataset}
\label{sec:analysis}
At the core, PanLex is a \emph{database}, served by a PostgreSQL database server, that ``represents \emph{assertions} about the \emph{meanings} of \emph{expressions}''\footnote{\url{http://panlex.org/tech/doc/design/panlex-db-design.pdf}}.
As of now, the database contains about 20 million meanings and 19 million expressions extracted from about 2,000 sources.

Assertions are factual claims attributed to a certain source and editor.
They do not represent absolute truths and as such it is valid that different editors make contradictory assertions based on the same sources.

@@

The lexical database of the PanLex project is based on different sources like monolingual dictionaries,
bilingual dictionaries, multilingual dictionaries, glossaries, standards, and thesauri.
The list of used sources can also be found on the web\footnote{\url{http://panlex.org/tech/plrefs.shtml}}.

@@

\begin{itemize}
  \item \emph{Expressions} are lemmas. For example, \emph{``go''} is a valid expression, whereas \emph{``went''} is not.
  Expressions are always given in a language variety and can only be given once per language variety.
  They were extracted from the above mentioned mono- or bilingual dictionaries or from other sources.
  \item \emph{Languages} in PanLex are identified using ISO 639-3\footnote{\url{http://www.sil.org/iso639-3/codes.asp}} individual and macrolanguage codes, ISO 639-2\footnote{\url{http://www.loc.gov/standards/iso639-2/}} collective codes and ISO 639-5\footnote{\url{http://www.loc.gov/standards/iso639-5/}} codes.
  \item \emph {Language Varieties} within a language are further distinguished by arbitrary unique integer identifiers.
  For example, ``eng'' is the ISO 639-3 code for the English language.
  Panlex defines various varieties of it, including ``English'' (eng-000), ``Simple English'' (eng-001) and ``British English'' (eng-005).
  Their labels, when possible, are autonyms, written in the native writing system.
  So, in contrast to mechanisms like IETF BCP 47\footnote{\url{http://tools.ietf.org/html/bcp47}} there is no need for a transcription. %  and one can use variety names like ``{زبان اوستایی}''.
  Additionally these variety identifiers and labels do not have to be registered like the IETF BCP 47 language subtags\footnote{\url{http://www.iana.org/assignments/language-subtag-registry}}.
  \item \emph{Meanings} of expressions are distinguished by editors based on their
  interpretation of an information source.
  Information sources can be dictionaries, thesauri, etc., and also personal knowledge. 
  The pair comprised of the editor and the information source is referred to as a meaning's \emph{approver}.
  Editors may add informal definitions to meanings in any language variety.
  \item \emph{Definitions} are optional descriptions of a meaning.
  They are given in a certain language variety.
  A description of the verb \emph{browse} for example could be \emph{``move or surf through various files on a computer, the Internet, etc.''}, marked as a definition in the ``English'' language variety.
  \item \emph{Denotations} relate expressions to \emph{meanings}.
  So homonyms are represented by one expression being connected to different meanings.
  Denotations may be enriched with part-of-speech tags selected from a closed list based on the
  Open Lexicon Interchange Format (OLIF) standard\footnote{\url{http://www.olif.net/}}.
  For example, \emph{fall} can be a verb or a noun for autumn.
  Furthermore, it is possible to generically tag denotations with metadata
  using key-value pairs.
  For instance, an English expression \emph{pig} when synonymous with
  \emph{police officer} could be annotated with \emph{pragmatics=vulgar}.
  \item \emph{Users} have editorial privileges over the language varieties and the approvers that they define.
  \item \emph{Licenses} are also considered by the PanLex project.
  At present there are ten different license categories an approver can be annotated with.
  They are \emph{public domain}, \emph{Creative Commons}, \emph{request} (meaning that one has to ask the author of the resource), \emph{GNU General Public License}, \emph{GNU Lesser General Public License}, \emph{GNU Free Documentation License}, \emph{MIT License}, \emph{copyright} (stating that there is a certain copyright holder), \emph{other} and \emph{unknown}.
\end{itemize}

An overview of the number of instances per entity in the current PanLex database is given in~\autoref{fig:plx-entity-counts}.
\begin{table}
\centering
\begin{scriptsize}
\begin{tabular}{lr}
\toprule
Entity & Instances \\
%\midrule
Expressions & 18,580,594 \\ % SELECT COUNT(ex) FROM ex;
%\midrule
Languages   & 7,839 \\ % SELECT COUNT(lc) FROM lc; -- all language codes (ISO 639-3, ISO 639-2, ISO 639-5) known in the PanLex db
% FIXME: there should not be more languages than language varieties
%\midrule
Language Varieties & 7,248 \\ % SELECT COUNT(lv) FROM lv;
%\midrule
Meanings    & 20,023,427 \\ % SELECT COUNT(mn) FROM mn;
%\midrule
Definitions & 2,522,605 \\ % SELECT COUNT(df) FROM df;
%\midrule
Denotations & 50,803,243 \\ % SELECT COUNT(dn) FROM dn;
%\midrule
Users       & 7 \\ % SELECT COUNT(us) FROM us;
%\midrule
Licenses    & 10 \\ % SELECT COUNT(li) FROM apli;
\bottomrule
\end{tabular}
\end{scriptsize}
\caption{Number of entities in the PanLex database}
\label{fig:plx-entity-counts}
\end{table}

@@

There are two noteworthy limitations of the database schema:
Approver combines a user with an information source, however the \emph{information source} is not modelled as a distinct entity.
Users are only associated with language variety and approver entities. However, whenever a given approver has more than one user with editorial privileges, it is not possible to track which denotations were added by which of the approver's authorized users.
\todo{What are the consequences? }

@@

The RDF analog of the PanLex \emph{meaning} is the \texttt{plx:Meaning}.
Entities of this class may have an identifier assigned with the \texttt{plx:identifier} property pointing to an \texttt{xsd:string} literal.
This meaning identifier represents ids given to the considered meaning in other data sources and may be used to build links to the meaning as represented on these sites.
Meanings may also have \emph{definitions}, entities of the \texttt{plx:Definition} class, giving a textual representation (\texttt{rdfs:label}) with more than just a lexeme in a certain language variety (\texttt{plx:languageVariety}).

@@

\subsection{Transformation workflow}
\label{sec:conversion}
\todo{how often in PanLex updated?}
The size of the PanLex database (including indexes) is approximately 18 GB,
which is big enough to make recurrent RDF conversion cumbersome:
A full RDF conversion bloats the dataset and results in data duplication.   
Using conventional hardware, it takes impractically long
to convert all the data. This makes
testing and debugging and fixing issues in the modelling or data conversion of the
data very time-consuming.
As the PanLex data already resides in a relational database, the use of a virtual
RDB 2 RDF mapping solution is a natural choice. The \emph{Sparqlify Mapping Language}
offers a very easy-to-use mapping language.
Essentially, these mappings consist of three clauses:
\begin{itemize}
  \item \emph{From} Specifies the logical SQL table (i.e. table, view or query) underlying the Sparqlify-ML view.
  \item \emph{With} Defines SPARQL variables by means of expressions over relational columns that yield RDF terms.
  \item \emph{Construct} The template (i.e. the set of triple patterns) to be generated in this view, based on the SPARQL variable definitions.
\end{itemize}
~\autoref{fig:ex:sparqlify-ml} shows an example of a Sparqlify-ML view for the
languages in PanLex: From each row of the \emph{i1} table three resources are
bound to the variable names \emph{?lang}, \emph{?iso3} and \emph{?lexvo3}.
Resources for \emph{?lang} become typed as a \emph{Language}in the PanLex and the \emph{schema.org} namespace.
\todo{Maybe we should link to lexvo directly.}
This view-based approach also demonstrates that changing the vocubulary or
adding support for new ones does not require an ETL process, and can therefore be done with little effort.
%PanLex's languages  iso 639-3 code 
%typed with plx:Language
%resources are created for
%the languages in PanLex, linked with \emph{lexvo}, and

@@

Since Lexvo covers all ISO 639-3 language codes, interlinking could be done with
the Sparqlify-ML view in Figure~\ref{fig:ex:sparqlify-ml}.
For DBpedia we were interested in creating \emph{valid} and
thus \emph{dereferenceable} links.
Therefore, we iterated the \emph{titles} datasets\footnote{\url{http://wiki.dbpedia.org/Downloads38} - Note that more languages are available in the raw download section at \url{http://downloads.dbpedia.org/3.8/}},
which map non-localized DBpedia URIs to their page titles in the respective language.
For each language version we normalized the labels: 
Unicode NFKD\footnote{\url{http://unicode.org/reports/tr15/}} normalization and
removal of punctuation characters were applied.
% PanLex now uses NFC, not NFKD, for character normalization and does not remove punctuation.
The DBpedia resources were then mapped to the PanLex expression that was equal to the resource's normalized label in the respective language.
Table ~\ref{fig:plx-dbp-link-counts} summarizes the number of links obtained.
\begin{table}
\centering
\begin{scriptsize}
\begin{tabular}{llll}
\toprule
Language & 639-1 & 639-3 & Links \\
\midrule
English    & en & eng & 1,415,241 \\
German     & de & deu &   224,146 \\
French     & fr & fra &   187,364 \\
Italian    & it & ita &   147,485 \\
Spanish    & sp & spa &   117,056 \\
Portuguese & pt & por &   112,266 \\
Polish     & pl & pol &   110,974 \\
Russian    & ru & rus &    68,040 \\
Czech      & cs & ces &    28,767 \\
Catalan    & ca & cat &    27,779 \\
Korean     & ko & kor &    24,912 \\
Turkish    & tr & tur &    22,258 \\
Bulgarian  & bg & bul &    19,431 \\
Hungarian  & hu & hun &    18,203 \\
Redo Slovene & sl & slv &  4,636 \\
Greek      & el & ell &  1,112 \\
\bottomrule
\end{tabular}
\end{scriptsize}
\caption{Number of DBpedia links per language}
\label{fig:plx-dbp-link-counts}
\end{table}

@@

With the RDF version of PanLex, a large lexical resource participates in the Web of Data,
making it available to a broader community. Users can easily explore the PanLex data using
the services provided by SNORQL and Pubby. This also increases the chance of development of
interesting mashups.
For example, the TeraDict translation service\footnote{\url{http://panlex.org/cgi-bin/apidemo-2.cgi}}
could be easily realized using simple SPARQL query as shown in
\todo{Create the query: translations in a given target language of an expression}. 
Anothe future usage scenario is to link PanLex to Wortschatz in order to 
improve the recall of word-by-word translations by considering co-occurences.

\section{Related Work}
\label{sec:related}
PanLex is an integration project of many existing lexical resources with
the goal of eventually covering all languages in the world, especially
ones having few speakers.
The extraction of information from linguistic sources, and
techniques for automatically inferring translations, are
relevant work discussed in \cite{panlex_probtrans}.

An important initiative is the
\emph{Global Wordnet Association}\footnote{\url{http://www.globalwordnet.org/}},
which offers a platform for sharing wordnets and defines several goals.
These include setting forth standards for uniformly representing wordnets
of different languages and establishing a universal index of meaning.

@@

At the level of the Semantic Web, there is currently a trend in publishing linguistic resources
as Linked Data\footnote{\url{http://www.w3.org/DesignIssues/LinkedData.html}} and via SPARQL{\footnote{\url{http://www.w3.org/TR/sparql11-query/}}}.
These efforts are referred to as the \emph{Linguistic Linked Open Data Cloud} (LLOD)\footnote{\url{http://linguistics.okfn.org/resources/llod/}}. 
%Some of these datasets already make use of quasi-standard linguistic Semantic Web Ontolgies, such as  
Several (quasi-)standard ontologies have been developed for covering different aspects of linguistic resources.
Examples include the \emph{Ontologies of Linguistic Annotation} (OLiA)~\cite{olia2010},
the \emph{Lexicon Model for Ontologies} (lemon)~\cite{lemon2011} for modeling lexicon and machine-readable dictionaries,
\emph{POWLA} for modeling linguistic corpora\cite{powla2012}
and the \emph{Natural Language Processing Interchange Format} (NIF)\footnote{\url{http://nlp2rdf.org/nif-1-0}}.

@@

\section{Conclusions and Future Work}
\label{sec:conclusion}
In this paper we describe the PanLex database and its conversion to RDF.
Based on our URI and vocabulary design, we created appropriate view definitions
for the virtual RDB to RDF solution \emph{Sparqlify} which carries out
the actual RDF transformation.
Furthermore, we interlinked the languages in PanLex with lexvo,
and created a total of \todo {number} derefenceable
links to DBpedia for expressions in 16 languages.
%We detailed the developed vocabulary and

In this work we also identified some weaknesses in
the orginal database design,
which we intend to overcome in the future:
The orginal dataset currently does not cleanly model information sources and approvers
as distinct entities.\todo{We should still map it like that with Sparqlify - resuse ap for the source}
Also, while it is possible to determine each meaning's and each denotation's approver, an approver can have multiple users with editorial privileges, so it is not possible to track each contributed, modified, and deleted meaning and denotation by editor.
Retaining this information seems beneficial for translation approches,
as this enables for example the attribution of qualities and relevances to particular editors.
